How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2021-08-01 00-author.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

## Creative Commons License {-}
How to write a PhD in Biological Sciences: a guide for the uninitiated is copyright to John Measey and distributed under a Creative Commons BY-NC-ND 4.0 International License. You may freely share this content for non-commercial purposes, with a source credit to [www.HowToWriteAPhD.org](www.HowToWriteAPhD.org).

@@

```{r iris}
knitr::kable(
  head(iris), caption = 'We must have the iris data and some R code, otherwise it would not seem right',
  booktabs = TRUE
)
```

>>>>

## Creative Commons License {-}
How to write a PhD in Biological Sciences: a guide for the uninitiated is copyright to John Measey and distributed under a [Creative Commons BY-NC-ND 4.0 International License](https://creativecommons.org/licenses/by-nc-nd/4.0/). You may freely share this content for non-commercial purposes, with a source credit to [www.HowToWriteAPhD.org](www.HowToWriteAPhD.org).

@@

### 2021-08-01 1.01-beforeyoustart.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

Right from the very start of your PhD, it is important for you to become familiar with the goal of what you are trying to achieve, how you plan to achieve it, and (believe it or not) what you plan to do afterwards. There are a lot of things to think about as you start your PhD, and for many of these it is far easier to make informed decisions from the outset. For example, reconciling yourself to using the [statistical package R](useR.html) will be much easier from the start, rather than only realising under pressure at the end of your PhD that there is no way that you can finish unless you take it on. Preparing yourself into the right mental framework will also be very important for your ability to work under pressure. Getting into good habits at the beginning will put you in a commanding position later. The aim of this first part of the book is to help you assess your current position, and realign yourself toward an optimum mindset.

@@

I prefer a thesis that fits together within an explicit [conceptual framework](PhDintro.html). I like PhD students to conceptualise how their chapters are interlinked, and to present this as a figure in the introduction of their [proposal](proposal.html). A useful concept to be aware of is the 'Hierarchy of Hypotheses' [@heger2018hierarchy]. In this approach, you are encouraged to consider the '[bigger question](bigidea.html)' in your research area, identifying both what studies have produced compelling evidence, and identifying knowledge gaps from a theory driven approach. Generally, it is only possible to consider these types of concepts once you are familiar with the literature (both theoretical and experimental).

>>>>

Right from the very start of your PhD, it is important for you to become familiar with the goal of what you are trying to achieve, how you plan to achieve it, and (believe it or not) what you plan to do afterwards. There are a lot of things to think about as you start your PhD, and for many of these it is far easier to make informed decisions from the outset. For example, reconciling yourself to using the [statistical package R](#useR) will be much easier from the start, rather than only realising under pressure at the end of your PhD that there is no way that you can finish unless you take it on. Preparing yourself into the right mental framework will also be very important for your ability to work under pressure. Getting into good habits at the beginning will put you in a commanding position later. The aim of this first part of the book is to help you assess your current position, and realign yourself toward an optimum mindset.

@@

I prefer a thesis that fits together within an explicit [conceptual framework](#PhDintro). I like PhD students to conceptualise how their chapters are interlinked, and to present this as a figure in the introduction of their [proposal](#proposal). A useful concept to be aware of is the 'Hierarchy of Hypotheses' [@heger2018hierarchy]. In this approach, you are encouraged to consider the '[bigger question](#bigidea)' in your research area, identifying both what studies have produced compelling evidence, and identifying knowledge gaps from a theory driven approach. Generally, it is only possible to consider these types of concepts once you are familiar with the literature (both theoretical and experimental).

### 2021-08-01 2.04-BigIdea.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

## So where would we find these big ideas?
There are quite a few papers that synthesise hypotheses in various areas of biology. Here I provide a few to get you started. Ask your advisor about relevant reviews of hypotheses in your area of biological sciences. For example, here are some theories and hypotheses in ecology [@vellend2010conceptual; @travassos-britto2021pragmatic; @catford2009reducing], while these wide ranging papers about so-called 'laws' in biology and ecology [@dhar2010laws; @lawton1999are]. Each of these papers will give you a list of big ideas, together with the citations for seminal papers that have built them. Clearly, you will need to find the papers that set out these big ideas in your own discipline. You will note that many of the theories are very old, with some dating back to Darwin. You can think of the way that such ideas are structured as a hierarchy of hypotheses (Figure \@ref(fig:JeschkeHagar61)). Building such a hierarchy of ideas in your own specialist subject area is good reference material to prompt [critical reading](critical.html) and reflection. You may want to take this structure forward when thinking how to [introduce your PhD chapters](PhDreview.html).

>>>>

## So where would we find these big ideas?
There are quite a few papers that synthesise hypotheses in various areas of biology. Here I provide a few to get you started. Ask your advisor about relevant reviews of hypotheses in your area of biological sciences. For example, here are some theories and hypotheses in ecology [@vellend2010conceptual; @travassos-britto2021pragmatic; @catford2009reducing], while these wide ranging papers about so-called 'laws' in biology and ecology [@dhar2010laws; @lawton1999are]. Each of these papers will give you a list of big ideas, together with the citations for seminal papers that have built them. Clearly, you will need to find the papers that set out these big ideas in your own discipline. You will note that many of the theories are very old, with some dating back to Darwin. You can think of the way that such ideas are structured as a hierarchy of hypotheses (Figure \@ref(fig:JeschkeHagar61)). Building such a hierarchy of ideas in your own specialist subject area is good reference material to prompt [critical reading](#critical) and reflection. You may want to take this structure forward when thinking how to [introduce your PhD chapters](#PhDreview).

### 2021-08-01 2.06-argument.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

There are other shorter forms that might suit you better. Consider another classic form of the scientific argument: “compare and contrast” which allows you as a writer to quickly familiarise your reader with some key examples. Another sentence structure to consider is explaining cause and effect. For practical wording examples of both of these classic arguments, [see part 2](phrasebank.html).

@@

**Absence of evidence is not evidence of absence** - The idea that positive evidence of the existence of a relationship is superior to a lack of the same evidence. This is also known as a Type II error (false negative), where the lack of finding a significant relationship does not lead to a conclusion that the relationship does not exist. Later we discuss how a [lack of finding of a significant result](transparency.html) leads to [confirmation bias](#TypeI) in science.

@@

Beware of copying an argument wholesale. You might well end up getting muddled, or worse perpetuating an error. Better go back and make sure that you understand the original premise and the works that promote or oppose this. Having told you to beware, I’m now going to encourage you to read, because [reading is one of the best ways in which you can learn about writing a logical argument](criticalreading.html).

@@

In this writing book, I concentrate on providing formulaic approaches to writing, because these are by far the easiest ways for inexperienced writers. However, I encourage you to learn and experiment with writing styles as you become more experienced. [Critical reading](criticalreading.html) is one of the best ways to learn about alternative writing styles: reading is probably the best way of improving your writing. Alternatively, and especially if the above hasn’t clicked with you, you can read more about writing a scientific argument (see for example [here](https://msu.edu/course/iah/231b/gifford/sciaargs.htm)). And there’s plenty more out there.

>>>>

There are other shorter forms that might suit you better. Consider another classic form of the scientific argument: “compare and contrast” which allows you as a writer to quickly familiarise your reader with some key examples. Another sentence structure to consider is explaining cause and effect. For practical wording examples of both of these classic arguments, [see part 2](#phrasebank).

@@

**Absence of evidence is not evidence of absence** - The idea that positive evidence of the existence of a relationship is superior to a lack of the same evidence. This is also known as a Type II error (false negative), where the lack of finding a significant relationship does not lead to a conclusion that the relationship does not exist. Later we discuss how a [lack of finding of a significant result](#transparency) leads to [confirmation bias](#TypeI) in science.

@@

Beware of copying an argument wholesale. You might well end up getting muddled, or worse perpetuating an error. Better go back and make sure that you understand the original premise and the works that promote or oppose this. Having told you to beware, I’m now going to encourage you to read, because [reading is one of the best ways in which you can learn about writing a logical argument](#critical).

@@

In this writing book, I concentrate on providing formulaic approaches to writing, because these are by far the easiest ways for inexperienced writers. However, I encourage you to learn and experiment with writing styles as you become more experienced. [Critical reading](#critical) is one of the best ways to learn about alternative writing styles: reading is probably the best way of improving your writing. Alternatively, and especially if the above hasn’t clicked with you, you can read more about writing a scientific argument (see for example [here](https://msu.edu/course/iah/231b/gifford/sciaargs.htm)). And there’s plenty more out there.

### 2021-08-01 2.07-storytelling.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

But the scientific paper that was written about this ‘story’ [@becker2018extreme] doesn’t have a narrative style, and doesn’t fit the description of a narrative that is discussed above. Instead, it follows the formula that is set out below: **Introduction, Materials and Methods, Results and Discussion**. This style does not treat the experiment as a complication on the way to the story’s goal. The structure introduces the rationale behind conducting the experiment, then objectively explains the findings, and lastly discusses their meaning in relation to what is already known. From this study, I can spin the story (above), because I know what the endpoint is. But this storyis just one interpretation of the correlation that was found. [Correlation is not causation](argument.html), and more studies would be needed to test the hypothesis concerning the underlying reasons suggested for the relationship observed. The key differences in the storytelling style from the scientific formula are the absence of a known goal at the start of the process (see a more detailed discussion of this by Yarden Katz, [-@katz2013storytelling] here).

There is a role for stories when communicating science as this increases interest, facilitates understanding, and enhances memory. This is particularly true when communicating science to the public and making it more accessible, but it also applies to interactions between scientists, for example, at conferences. The [presentations](#proposal-presentation.html) that tell a story and entertain are those we tend to remember. Not easy, but if we do want to communicate well with each other, then we need to learn the art of storytelling, without compromising our scientific objectivity. In addition, there are lots of lessons from storytelling that we can use within scientific papers, and to which I subscribe wholeheartedly. For example, don’t be boring, don’t [bamboozle](style.html), write an active narrative (not in the passive tense), use the magic of threes, and many more.

>>>>

But the scientific paper that was written about this ‘story’ [@becker2018extreme] doesn’t have a narrative style, and doesn’t fit the description of a narrative that is discussed above. Instead, it follows the formula that is set out below: **Introduction, Materials and Methods, Results and Discussion**. This style does not treat the experiment as a complication on the way to the story’s goal. The structure introduces the rationale behind conducting the experiment, then objectively explains the findings, and lastly discusses their meaning in relation to what is already known. From this study, I can spin the story (above), because I know what the endpoint is. But this storyis just one interpretation of the correlation that was found. [Correlation is not causation](#argument), and more studies would be needed to test the hypothesis concerning the underlying reasons suggested for the relationship observed. The key differences in the storytelling style from the scientific formula are the absence of a known goal at the start of the process (see a more detailed discussion of this by Yarden Katz, [-@katz2013storytelling] here).

There is a role for stories when communicating science as this increases interest, facilitates understanding, and enhances memory. This is particularly true when communicating science to the public and making it more accessible, but it also applies to interactions between scientists, for example, at conferences. The [presentations](#proposal-presentation) that tell a story and entertain are those we tend to remember. Not easy, but if we do want to communicate well with each other, then we need to learn the art of storytelling, without compromising our scientific objectivity. In addition, there are lots of lessons from storytelling that we can use within scientific papers, and to which I subscribe wholeheartedly. For example, don’t be boring, don’t [bamboozle](#style), write an active narrative (not in the passive tense), use the magic of threes, and many more.

### 2021-08-01 2.08-citations.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

Within a manuscript for a scientific journal or in your thesis, you can expect that your [introduction](introduction.html) and [discussion](discussion.html) sections are going to be full of citations. The methodology will likely also be citation rich. By using a lot of citations in your methods, you can often save your needing to write a lot of sections which would otherwise be very detailed.

Later in the book ([Writing with a Formula](formula.html)), I explain how to start writing with an outline, then flesh out the outline with citations from your reading, and plan which citations you are going to use carefully before you start. This won't stop you from adding more later, but this ensures that your work is built on your scholarly endeavour.

@@

Sometimes primacy is less important, especially if the concept is well known and/or has changed substantially. Then you should cite the most recent work that shows the ideas that you are wanting to show. Other times, primacy is more important - especially when few have built on ideas or concepts since they were put forward, or in the [taxonomic literature](taxonomy.html). If in doubt, place a citation to the original study and the most recent study, then flag this using a comment for your advisor to consider. The important point here is for you to give credit where it's due and not to overlook those who put in the hard work to publish original content.

Some have argued that the rule of priority is of benefit to science as it drives competition and this spurs creativity. On the other hand, the winner-takes-all approach of being first with something drives secrecy and dishonesty, and does not genuinely reflect the team effort that is the [science project](lifescientific.html) [@casadevall2012reforming].

@@

Obviously, everything you cite needs to be in the References (or [Literature Cited](references.html)) section, and you may well need to spend time deleting extra stuff. You can get around this common issue by using a citation manager like [Mendeley, Zotero or EndNote](references.html). I’d always suggest that you use one of these tools as they can really help with your reading too. These days they are busy turning into a kind of scientific social network. They regularly make suggestions of what you could be reading based on what you read. This can be useful.

>>>>

Within a manuscript for a scientific journal or in your thesis, you can expect that your [introduction](#introduction) and [discussion](#discussion) sections are going to be full of citations. The methodology will likely also be citation rich. By using a lot of citations in your methods, you can often save your needing to write a lot of sections which would otherwise be very detailed.

Later in the book ([Writing with a Formula](#formula)), I explain how to start writing with an outline, then flesh out the outline with citations from your reading, and plan which citations you are going to use carefully before you start. This won't stop you from adding more later, but this ensures that your work is built on your scholarly endeavour.

@@

Sometimes primacy is less important, especially if the concept is well known and/or has changed substantially. Then you should cite the most recent work that shows the ideas that you are wanting to show. Other times, primacy is more important - especially when few have built on ideas or concepts since they were put forward, or in the [taxonomic literature](#taxonomy). If in doubt, place a citation to the original study and the most recent study, then flag this using a comment for your advisor to consider. The important point here is for you to give credit where it's due and not to overlook those who put in the hard work to publish original content.

Some have argued that the rule of priority is of benefit to science as it drives competition and this spurs creativity. On the other hand, the winner-takes-all approach of being first with something drives secrecy and dishonesty, and does not genuinely reflect the team effort that is the [science project](#lifescientific) [@casadevall2012reforming].

@@

Obviously, everything you cite needs to be in the References (or [Literature Cited](#references)) section, and you may well need to spend time deleting extra stuff. You can get around this common issue by using a citation manager like [Mendeley, Zotero or EndNote](#references). I’d always suggest that you use one of these tools as they can really help with your reading too. These days they are busy turning into a kind of scientific social network. They regularly make suggestions of what you could be reading based on what you read. This can be useful.

### 2021-08-01 2.09-databases.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

There are two typical ways in which you can search for literature about your area of interest. The first is to use keywords and the second is to use citations of key references. I suggest that you use both of these tactics when doing your initial literature search. Further, I’d suggest that it’s worth your time keeping a record of what you have searched for, how and when and then download the results. This is called a systematic literature review. Using this tactic from the outset of your interactions with literature databases will give you a good idea of what proportion of the literature you are reading, and you might then be able to convert your finding into an objective literature review - making a great first [introductory chapter](PhDintro.html) for your thesis.

@@

While searching for keywords sounds easy, picking the keywords that you should use is more difficult and requires a lot of practice with the databases and your key references. Remember that keywords are often not single words, but also include short phrases. Look at all of the key references and the keywords that they contain. Any keywords that they share are going to be important for you. Make a note of all relevant keywords for your searches, as ultimately, this will help you when you need to determine the keywords for [your own chapters](title.html).

@@

An invaluable way in which databases have become extremely useful is the ability to move items that you have found during your search directly into your [reference manager](referencesoftware.html). Beware that these regularly come with errors, and that the amount of errors is related to your database source (Scopus and WoS are much better than Google Scholar). This means that after importing the references, you will need to go back to your reference manager and edit them or they will come out incorrectly when you use them. Of particular note to biologists is the need to have all species names in [italics](taxonomy.html), and you'll need to do this yourself.

@@

The [ORCID](orcid.html) number has been erected to provide a common platform for [authors to curate](title.html), and now some journals and funding authorities won't allow you to submit or apply without one.

>>>>

There are two typical ways in which you can search for literature about your area of interest. The first is to use keywords and the second is to use citations of key references. I suggest that you use both of these tactics when doing your initial literature search. Further, I’d suggest that it’s worth your time keeping a record of what you have searched for, how and when and then download the results. This is called a systematic literature review. Using this tactic from the outset of your interactions with literature databases will give you a good idea of what proportion of the literature you are reading, and you might then be able to convert your finding into an objective literature review - making a great first [introductory chapter](#PhDintro) for your thesis.

@@

While searching for keywords sounds easy, picking the keywords that you should use is more difficult and requires a lot of practice with the databases and your key references. Remember that keywords are often not single words, but also include short phrases. Look at all of the key references and the keywords that they contain. Any keywords that they share are going to be important for you. Make a note of all relevant keywords for your searches, as ultimately, this will help you when you need to determine the keywords for [your own chapters](#title).

@@

An invaluable way in which databases have become extremely useful is the ability to move items that you have found during your search directly into your [reference manager](#referencesoftware). Beware that these regularly come with errors, and that the amount of errors is related to your database source (Scopus and WoS are much better than Google Scholar). This means that after importing the references, you will need to go back to your reference manager and edit them or they will come out incorrectly when you use them. Of particular note to biologists is the need to have all species names in [italics](#taxonomy), and you'll need to do this yourself.

@@

The [ORCID](#orcid) number has been erected to provide a common platform for [authors to curate](#title), and now some journals and funding authorities won't allow you to submit or apply without one.

### 2021-08-01 2.10-referencesortware.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

Because published papers play such a fundamental role in your [reading and comprehension](critical.html) of a subject, and [citations](citations.html) are pivotal in chapter or paper writing, putting these elements together in a good reference manager makes a lot of sense. A reference manager is a piece of software in which you can store and view all of your papers (in pdf format or via links), make notes on them, and use them to make citations and build bibliographies. Unsurprisingly, they become something that you may well use a lot during the course of your PhD, and it is worth thinking carefully about which one you want to use.

@@

My advice here is (like software for [writing your PhD](software.html)) to pick the format that your references are going to be stored in carefully. For example, the BibTex format (see below) is a generic format that many of the software packages are capable of importing and exporting. Using a generic referencing format means that you will (in future) be able to change packages with minimum loss. Note that there is a difference between being able to output in BibTex and storage as BibTex files. In the case of the former, you will require a working piece of software in order to output from that format. If your software stores references as BibTex files, you are always able to take this file and use it elsewhere.

@@

 - Does it integrate with your [word processing package](software.html)?
 - Does your university have a licence?
 - Do other members of your lab use this software?
   + is there a communal repository?
 - Can you use it to read files and make notes on them?
 - Does it output into referencing formats that will be useful to you?
   + Or will it allow you to make a custom output format?
 - How much time will it take you to learn to use it?
 - Is there good support available?

>>>>

Because published papers play such a fundamental role in your [reading and comprehension](#critical) of a subject, and [citations](#citations) are pivotal in chapter or paper writing, putting these elements together in a good reference manager makes a lot of sense. A reference manager is a piece of software in which you can store and view all of your papers (in pdf format or via links), make notes on them, and use them to make citations and build bibliographies. Unsurprisingly, they become something that you may well use a lot during the course of your PhD, and it is worth thinking carefully about which one you want to use.

@@

My advice here is (like software for [writing your PhD](#software)) to pick the format that your references are going to be stored in carefully. For example, the BibTex format (see below) is a generic format that many of the software packages are capable of importing and exporting. Using a generic referencing format means that you will (in future) be able to change packages with minimum loss. Note that there is a difference between being able to output in BibTex and storage as BibTex files. In the case of the former, you will require a working piece of software in order to output from that format. If your software stores references as BibTex files, you are always able to take this file and use it elsewhere.

@@

 - Does it integrate with your [word processing package](#software)?
 - Does your university have a licence?
 - Do other members of your lab use this software?
   + is there a communal repository?
 - Can you use it to read files and make notes on them?
 - Does it output into referencing formats that will be useful to you?
   + Or will it allow you to make a custom output format?
 - How much time will it take you to learn to use it?
 - Is there good support available?

### 2021-08-01 2.10a-sortware.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

Another important consideration is whether or not your chosen word processing package will allow seamless integration with your [reference manager](referencesoftware.html). Your lab may have a communal repository of references, and so this may then constrain what you are comfortable using.

If you feel comfortable with a word processing package with a nice GUI, like Microsoft Word, then go ahead and use it. These certainly have advantages like being able to get a synonym at the click of a mouse (Figure \@ref(fig:synonym)). My best advice is that you do not save your files in their default ".docx" format. Instead, make sure that all of your files are saved as .rtf, .txt or .html. This should make little difference to your opening and editing but will mean that all of your texts are forward compatible with other kinds of software. It will reduce your file size and reduce the chances of file corruption: i.e. when part of the file malfunctions and you lose the contents. That’s right, in addition to loss of data by back up deficiencies [see Part III](data.html), you also have the risk of having corrupt files that will also mean you lose all of the contents. Even if you always use Word and never use any other file format, do yourself a favour and save each version as a .rtf or .txt file as well. Just like a back up, this means that you won’t have lost all of the text if a file goes corrupt. And yes, I do know people who have multiple backup versions of corrupt files: it’s happened to me.

@@

The real problem with word processing packages like Microsoft Word is when files get very large as they tend to do during a PhD. Once you start putting chapters together especially if they include embedded images tables and the like you will find that many word processors start misbehaving. This can be especially frustrating in the [closing phases of your PhD](part4.html) when you're trying to submit it. Especially if there's a tight deadline. However this problem is easily overcome by chunking your thesis into documents equivalent to chapters and then merging PDFs from each chapter.

@@

I would suggest keeping at least one older version of each file to minimise the possibility of losing the content through corrupt files, but you should also be keeping those .rtf or .txt files as well, just in case. Because of my early experience with file corruption, I started to keep all versions of files. This is unnecessary, and I’ve never needed to go back to anything but the last version. Nevertheless, text files take up very little space and it’s not a problem to keep more versions if you want to. More on this in the chapter on [data management](data.html).

>>>>

Another important consideration is whether or not your chosen word processing package will allow seamless integration with your [reference manager](#referencesoftware). Your lab may have a communal repository of references, and so this may then constrain what you are comfortable using.

If you feel comfortable with a word processing package with a nice GUI, like Microsoft Word, then go ahead and use it. These certainly have advantages like being able to get a synonym at the click of a mouse (Figure \@ref(fig:synonym)). My best advice is that you do not save your files in their default ".docx" format. Instead, make sure that all of your files are saved as .rtf, .txt or .html. This should make little difference to your opening and editing but will mean that all of your texts are forward compatible with other kinds of software. It will reduce your file size and reduce the chances of file corruption: i.e. when part of the file malfunctions and you lose the contents. That’s right, in addition to loss of data by back up deficiencies [see Part III](#data), you also have the risk of having corrupt files that will also mean you lose all of the contents. Even if you always use Word and never use any other file format, do yourself a favour and save each version as a .rtf or .txt file as well. Just like a back up, this means that you won’t have lost all of the text if a file goes corrupt. And yes, I do know people who have multiple backup versions of corrupt files: it’s happened to me.

@@

The real problem with word processing packages like Microsoft Word is when files get very large as they tend to do during a PhD. Once you start putting chapters together especially if they include embedded images tables and the like you will find that many word processors start misbehaving. This can be especially frustrating in the [closing phases of your PhD](#part4) when you're trying to submit it. Especially if there's a tight deadline. However this problem is easily overcome by chunking your thesis into documents equivalent to chapters and then merging PDFs from each chapter.

@@

I would suggest keeping at least one older version of each file to minimise the possibility of losing the content through corrupt files, but you should also be keeping those .rtf or .txt files as well, just in case. Because of my early experience with file corruption, I started to keep all versions of files. This is unnecessary, and I’ve never needed to go back to anything but the last version. Nevertheless, text files take up very little space and it’s not a problem to keep more versions if you want to. More on this in the chapter on [data management](#data).

### 2021-08-01 2.15-concise.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

In this book, I have put an emphasis on a [writing formula](formula.html) in order to get started writing. Many students have a problem in knowing where and how to get started with what seems to be a daunting prospect: writing a proposal or writing the entire thesis. I still feel that getting started quickly and efficiently with some confidence is one of your most important first steps. But once you've made it through your first draft, you'll begin to find that you need to start refining your writing. That is saying what you need to say as much in less words and space. A bit like constantly rewriting your Tweets to get the same information into 140 characters (yes, it's been increased to 280 but that's not as concise).

@@

Although we talk about writing science here, the grammatical rules of writing (in English) are fairly universal (with some local exceptions), so do practice your writing. Maybe write a letter (or long email) instead of making a call. It'll also help you to [read critically](critical.html). If you are a keen reader, then you could do worse than reading some Kurt Vonnegut [-@vonnegut1980how].

@@

It's important to avoid distracting your reader (see [Part II](part2.html)). Your aim is to be thorough, which will mean including all relevant information, but don't allow your writing to sideline them into taking the wrong direction. You need to stick to your narrative like a highway. The highway analogy is useful as although on a highway you will see signs to other places that you can point out to the reader, you shouldn't turn off the highway to get to your destination. The best of highways will also be free of traffic jams (think long, complex sentences here), and toll free ([no paywall](paywall.html)).

@@

Also, please don't cut and paste sections of [methods](methods.html) from one part of your thesis to another. Just don't do it.

People also have a tendency to build the [abstract](abstract.html) by copying and pasting text from the main sections. Don't do it. The reader will quickly see the repetitions and become bored. Similarly, the [conclusion/summary](discussion.html) section is also often copied from the lines above. If it's not worth writing again, then it wasn't worth writing before. Make it fresh, and keep it interesting!

@@

Here's a rule that Scott didn't have. Perhaps it should have been inserted higher up the list. It would help you to have your work read through by one of your colleagues before you [give it to your advisor](goodenough.html). But above all, you must be prepared to read __all of your work__ yourself, and be prepared to work on it when you read. Edit and improve it on every read. Not just the first time you write it, but for **every** version. Please don't ever expect your advisor to read something through when you can't be bothered to read it yourself.

Being concise is a great way to write well, and the [abstract](abstract.html) is a case in point. Please don't forget to read Scott's 10 rules as written by Scott [@hotaling2020simple].

>>>>

In this book, I have put an emphasis on a [writing formula](#formula) in order to get started writing. Many students have a problem in knowing where and how to get started with what seems to be a daunting prospect: writing a proposal or writing the entire thesis. I still feel that getting started quickly and efficiently with some confidence is one of your most important first steps. But once you've made it through your first draft, you'll begin to find that you need to start refining your writing. That is saying what you need to say as much in less words and space. A bit like constantly rewriting your Tweets to get the same information into 140 characters (yes, it's been increased to 280 but that's not as concise).

@@

Although we talk about writing science here, the grammatical rules of writing (in English) are fairly universal (with some local exceptions), so do practice your writing. Maybe write a letter (or long email) instead of making a call. It'll also help you to [read critically](#critical). If you are a keen reader, then you could do worse than reading some Kurt Vonnegut [-@vonnegut1980how].

@@

It's important to avoid distracting your reader (see [Part II](#part2)). Your aim is to be thorough, which will mean including all relevant information, but don't allow your writing to sideline them into taking the wrong direction. You need to stick to your narrative like a highway. The highway analogy is useful as although on a highway you will see signs to other places that you can point out to the reader, you shouldn't turn off the highway to get to your destination. The best of highways will also be free of traffic jams (think long, complex sentences here), and toll free ([no paywall](#paywall)).

@@

Also, please don't cut and paste sections of [methods](#methods) from one part of your thesis to another. Just don't do it.

People also have a tendency to build the [abstract](#abstract) by copying and pasting text from the main sections. Don't do it. The reader will quickly see the repetitions and become bored. Similarly, the [conclusion/summary](#discussion) section is also often copied from the lines above. If it's not worth writing again, then it wasn't worth writing before. Make it fresh, and keep it interesting!

@@

Here's a rule that Scott didn't have. Perhaps it should have been inserted higher up the list. It would help you to have your work read through by one of your colleagues before you [give it to your advisor](#goodenough). But above all, you must be prepared to read __all of your work__ yourself, and be prepared to work on it when you read. Edit and improve it on every read. Not just the first time you write it, but for **every** version. Please don't ever expect your advisor to read something through when you can't be bothered to read it yourself.

Being concise is a great way to write well, and the [abstract](#abstract) is a case in point. Please don't forget to read Scott's 10 rules as written by Scott [@hotaling2020simple].

### 2021-08-01 2.20-proposal.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

Your final thesis will be a set of papers (data chapters sandwiched between an [introduction and discussion](PhDintro.html)), and so prepare the proposal in the same way. That is, you will normally have two (MSc) or five (PhD) data chapters. In the proposal, each chapter will consist of three parts.

@@

I’m not going to go into what’s inside the introduction here, as it’s already been done in other chapters on formulaic writing (see later in this [part](formula.html)) and specifically about the introduction (see [part 3](introduction.html)). Although you may feel that there's no point putting a lot of effort into this introduction it serves as a literature review for your upcoming PhD study, and is well worth the time and effort.

@@

What’s in the [Methods and Materials](methods.html) section is also covered here. Essentially you need to:

- Introduce important aspects of your study organism and/or study site
- Describe exactly how you’ll set up your experiment and/or collect your samples
- You must explicitly state how each of the variables (introduced in the [introduction](introduction.html)) is collected
- (If necessary) Describe how you’ll turn your collected data into the variables needed for your analyses
- Data analyses (how you’ll test your [hypotheses](hypothesis.html)). This should include a description of the statistical or analytical techniques. It’s useful to structure this section by each of the questions/hypotheses that you are posing

@@

Writing grant proposals is essentially the same as your thesis proposal, and so you can use this document (with some modifications to tailor it to the grant objectives) to [raise money](funding.html) for your work.

@@

This is the subject of another chapter (see [part 3](transparency.html)).

>>>>

Your final thesis will be a set of papers (data chapters sandwiched between an [introduction and discussion](#PhDintro)), and so prepare the proposal in the same way. That is, you will normally have two (MSc) or five (PhD) data chapters. In the proposal, each chapter will consist of three parts.

@@

I’m not going to go into what’s inside the introduction here, as it’s already been done in other chapters on formulaic writing (see later in this [part](#formula)) and specifically about the introduction (see [part 3](#introduction)). Although you may feel that there's no point putting a lot of effort into this introduction it serves as a literature review for your upcoming PhD study, and is well worth the time and effort.

@@

What’s in the [Methods and Materials](#methods) section is also covered here. Essentially you need to:

- Introduce important aspects of your study organism and/or study site
- Describe exactly how you’ll set up your experiment and/or collect your samples
- You must explicitly state how each of the variables (introduced in the [introduction](#introduction)) is collected
- (If necessary) Describe how you’ll turn your collected data into the variables needed for your analyses
- Data analyses (how you’ll test your [hypotheses](#hypothesis)). This should include a description of the statistical or analytical techniques. It’s useful to structure this section by each of the questions/hypotheses that you are posing

@@

Writing grant proposals is essentially the same as your thesis proposal, and so you can use this document (with some modifications to tailor it to the grant objectives) to [raise money](#funding) for your work.

@@

This is the subject of another chapter (see [part 3](#transparency)).

### 2021-08-01 2.21-proposal-presentation.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

Despite previous sections [of this book](storytelling.html) cautioning you against storytelling, you really need to have turn your study into a story in order to present it well. This means that you know what the message you want to reveal to the audience is.  When you develop your story, be imaginative and creative. There are plenty of different styles to telling stories, and you should feel free to borrow styles that best fit your scenario. For example, you can reveal the punch line right at the beginning, and then work back through the presentation to show how you arrive there (but always try to keep something to reveal on the way there). Props are also excellent to produce in talks, something that you can bring with you, and show to an audience, that reveals more about your study. You want your audience to understand your story, but you also want them to remember it. Anything that will help them remember is likely to be a good prop.

>>>>

Despite previous sections [of this book](#storytelling) cautioning you against storytelling, you really need to have turn your study into a story in order to present it well. This means that you know what the message you want to reveal to the audience is.  When you develop your story, be imaginative and creative. There are plenty of different styles to telling stories, and you should feel free to borrow styles that best fit your scenario. For example, you can reveal the punch line right at the beginning, and then work back through the presentation to show how you arrive there (but always try to keep something to reveal on the way there). Props are also excellent to produce in talks, something that you can bring with you, and show to an audience, that reveals more about your study. You want your audience to understand your story, but you also want them to remember it. Anything that will help them remember is likely to be a good prop.

### 2021-08-01 2.22-transparency.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

It is probably too early to tell [@rubin2020does], but it is certainly a good place to start [@nosek2018preregistration]. The more researchers know and subscribe to [transparency](transparency2.html) in their research, the more it will shift the culture in science for the better [@forstmeier2017detecting]. However, there are more biases existing in science, and the system needs buy-in from stakeholders to promote equity by elimination of bias.

>>>>

It is probably too early to tell [@rubin2020does], but it is certainly a good place to start [@nosek2018preregistration]. The more researchers know and subscribe to transparency in their research, the more it will shift the culture in science for the better [@forstmeier2017detecting]. However, there are more biases existing in science, and the system needs buy-in from stakeholders to promote equity by elimination of bias.

### 2021-08-01 3.10-Figures.Rmd
Add files via upload
# corrections after July 2021 Copy Edits

I encourage my students to draw conceptual diagrams of how the thesis chapters relate to one another (Figure \@ref(fig:thesis-construction)), and then to include this diagram in the introduction to the thesis to help examiners and any other readers to understand how the chapters are interrelated (see [Part IV](PhDintro.html)).

>>>>

I encourage my students to draw conceptual diagrams of how the thesis chapters relate to one another (Figure \@ref(fig:thesis-construction)), and then to include this diagram in the introduction to the thesis to help examiners and any other readers to understand how the chapters are interrelated (see [Part IV](#PhDintro)).
